Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-04-02 document/ms.tex
smoothness prior

% Document specific variables.
\newcommand{\rate}{\ensuremath{\eta}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\theta}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\entry}{\ensuremath{S}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}

\newcommand{\interim}{{\ensuremath{\bvec{\alpha}}}}

@@

\dfmtodo{Discuss binned model and choice of priors.}

\dfmtodo{Show where the normalization comes in and describe how it acts
qualitatively.}

@@

>>>>

% Document specific variables.
\newcommand{\rate}{\ensuremath{\eta}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\theta}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\entry}{\ensuremath{S}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}
\newcommand{\binheight}{{\ensuremath{w}}}
\newcommand{\binheights}{{\ensuremath{\bvec{\binheight}}}}

\newcommand{\smoothreg}{{\ensuremath{\lambda}}}
\newcommand{\interim}{{\ensuremath{\bvec{\alpha}}}}

@@

In the following sections, we will model the occurrence rate
$\rate(\period,\,\radius)$ as a two dimensional grid with $N_\period$ bins in
$\ln\period$ and $N_\radius$ bins in $\ln\radius$.
We choose to fix the bin locations leaving only the bin heights
$\binheights = \{\binheight_{ij}\}_{i=1,j=1}^{N_\period,N_\radius}$ as
free parameters.
In practice, the results seem to be more numerically stable---but the model is
exactly equivalent---when we model the rate as
\begin{eqnarray}
\rate(\period,\,\radius) &=& \rate_0\,p(\period,\,\radius)
\end{eqnarray}
where $\rate_0$ is the total rate of planets in the sample and the relative
probability factor is subject to the constraint
\begin{eqnarray}
1 &=& \int p(\period,\,\radius) \dd\period\dd\radius\quad.
\end{eqnarray}
In this model, the parameter set \ratepars\ includes $\rate_0$ and the heights
of all but one of the bins in $p(\period,\,\radius)$.

Since the choice of bin locations and number is arbitrary, the most
conservative strategy is to choose very small bins---yielding an effectively
non-parametric model---and choose a prior (or regularization) that encodes any
prior expectations about the distribution.
In particular, it seems reasonable to expect the distribution to be
``smooth''.
The most extreme version of this model would be a completely non-parametric
rate function---generated by, for example, a Gaussian process
(\citealt{poiss-gp})---but this discussion is beyond the scope of this
\paper.
Instead, we'll simply choose small bins and add a term to the likelihood
function that penalizes large changes between neighboring bins.
Specifically, we choose a term of the form
\begin{eqnarray}
\ln p(\binheights\,|\,\smoothreg_\period,\,\smoothreg_\radius) &=&
-\frac{1}{2}\sum_{i=2}^{N_\period}
\frac{|\binheights_{i,\cdot} - \binheights_{i-1,\cdot}|}{\smoothreg_\period}
-\frac{1}{2}\sum_{j=2}^{N_\radius}
\frac{|\binheights_{\cdot,j} - \binheights_{\cdot,j-1}|}{\smoothreg_\radius}
-\frac{1}{2}\ln (\smoothreg_\period\,\smoothreg_\radius)
\quad.
\end{eqnarray}
In our final analysis, we'll marginalize over the regularization strengths
$\smoothreg_\period$ and $\smoothreg_\radius$ but we'll also explore the
effects of this choice below.

@@

\bibitem[Adams, Murray \& MacKay(2009)]{poiss-gp}
Adams, R.~P., Murray, I., \& MacKay, D.~J.~C.\ 2009, ICML, 2009, 9
(\href{http://homepages.inf.ed.ac.uk/imurray2/pub/09poisson/}{online})
